---
title: "EBMF Fit to ROSMAP Data (Greedy Fit, 100 Factors)"
author: "Jason Willwerscheid"
date: "2023-12-04"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I run `flashier` with [group priors](group_priors.html) and `greedy.Kmax = 100`. I wanted to see whether the full 100 factors would be added (they are!), and if so, how they are contributing to the overall fit (in terms of PVE and sparsity).

The code below has been pre-run; results are loaded here:

```{r output, message=FALSE, warning=FALSE}
library(tidyverse)
library(flashier)

load("./output/greedy.RData")
```

## Pre-process Data

As before, I load in the ROSMAP data and remove sites with at least 50% missing data across subjects. Additionally, **I transform the data using a probit transformation** (following discussion with Gao Wang and Alexandre Pelletier).

```{r eval=FALSE}
dat <- qnorm(dat)
```


## Fit EBMF

The `flashier` code appears as follows. **I impute missing data using `softImpute`** (following discussion with Gao Wang) and greedily add 100 factors without backfitting.

```{r eval=FALSE}
which_na <- is.na(dat)

t0 <- Sys.time()
si_fit <- softImpute(dat) # Default is rank 2.
t_si <- Sys.time() - t0

dat <- complete(dat, si_fit)

# Greedy fit.
t0 <- Sys.time()
fl <- flash_init(dat, var_type = 0) |>
  flash_set_verbose(3) |>
  flash_greedy(Kmax = 100L,
               ebnm_fn = c(flash_ebnm(group = sites_to_keep$chrom), 
                           ebnm_normal)) 
tg <- Sys.time() - t0
```


## Fitting Times

The addition of 100 greedy factors required:

```{r tg}
tg
```

This is about the same as was required to fit 50 greedy factors *with* backfitting. And in fact I would expect the greedy algorithm to scale quadratically in $K$ (since fitting the $K$th factor requires doing matrix operations with matrices of size $n \times (K - 1)$ and $p \times (K - 1)$).


## PVE of Factors

After $k = 25$ or so, there is no clear drop-off in the scree plot:

```{r pve}
plot(fl, include_scree = TRUE, include_pm = FALSE)
```


## Sparsity

The sparsity heatmap tells a similar story to before. The X chromosome is again estimated much differently from the others, with differences among other chromosomes remaining small. 

It is worth noting that **most factors after $k = 50$ or so are very sparse.** These might be primarily fitting noise (especially since the PVE is so small).

```{r sparsity}
pi0 <- t(sapply(fl$L_ghat, function(x) sapply(x, function(y) y$pi[1])))
tib <- as_tibble(pi0) |>
  mutate(Factor = 1:nrow(pi0)) |>
  pivot_longer(-Factor, names_to = "Chromosome", values_to = "Sparsity") |>
  mutate(Chromosome = str_remove(Chromosome, "chr")) |>
  mutate(Chromosome = factor(Chromosome, levels = c(as.character(1:22), "X")))
ggplot(tib, aes(x = Chromosome, y = Factor, fill = Sparsity)) + 
  geom_tile() +
  scale_fill_gradient(low = "black", high = "white") +
  scale_y_reverse(breaks = seq(5, 50, by = 5)) +
  theme_minimal()
```


## Conclusions

I am not sure this gives much guidance about which value of $K$ we should choose. From the scree plot it appears that 40--50 factors should suffice, and the sparsity plot lends some support to this choice. 

(I tried running `PCAtools::chooseMarchenkoPastur` as well, but the MP limit is well below the variance explained by the 100th factor. **I suspect that unless we have a good way to estimate the (theoretical) variance of the random noise the MP limit will not be usable.**)
