---
title: "Estimating K for the ROSMAP dataset via Parallel Analysis"
author: "Jason Willwerscheid"
date: "2023-12-15"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I try a parallel analysis approach to estimate $K$ for the ROSMAP data.

Again I pre-run the code; results are loaded here:

```{r output, message=FALSE, warning=FALSE}
library(tidyverse)
library(flashier)

load("./output/paranalysis10factors.RData")
load("./output/paranalysis10iters.RData")
```

## Pre-process Data

As in previous write-ups, I load in the ROSMAP data, remove sites with at least 50% missing data across subjects, transform the data using a probit transformation, and impute missing data using `softImpute`.


## Permute columns and fit EBMF

In a first pass, I fit 10 factors to see how the PVE decays.

```{r eval=FALSE}
set.seed(666)
pdat <- t(apply(dat, 1, sample))

# Greedy fit.
t0 <- Sys.time()
fl <- flash_init(pdat, var_type = 0) |>
  flash_set_verbose(2) |>
  flash_greedy(Kmax = 10L,
               ebnm_fn = c(flash_ebnm(group = sites_to_keep$chrom), 
                           ebnm_normal)) 
t10 <- Sys.time() - t0
```

Fitting time:

```{r}
t10
```

Scree plot:

```{r}
plot(fl, include_pm = FALSE, include_scree = TRUE)
```

In a parallel analysis style approach we might use the PVE of the second factor as a cutoff. If we refer back to our 100-factor greedy fit we find that this would set $K$ around 30:

```{r}
cutoff <- fl$pve[2]
load("./output/greedy.RData")
plot(fl, include_pm = FALSE, include_scree = TRUE) + 
  geom_hline(aes(yintercept = cutoff), linetype = "dashed")
```


## PVE or ELBO?

There is at present no way to stop adding factors when they drop below a given PVE. How can we implement this permutation-based approach in general, when we do not have any idea about what range of $K$ we should be looking for? Perhaps we can check the PVE after every 5--10 factors added and stop when we find $\kappa$ factors beneath the threshold. 

Another possibility is to use the difference in ELBO as a criterion (i.e., look at improvements in ELBO yielded by the addition of a second factor to the permuted datasets).

Here I do ten separate permutations of the ROSMAP dataset, fit two factors to each, and track both the PVE of, and the improvement in ELBO yielded by, the second factor:

```{r, eval=FALSE}
niter <- 10
tlist <- numeric(niter)
pvelist <- numeric(niter)
elbodifflist <- numeric(niter)

set.seed(999)
for (i in 1:niter) {
  t0 <- Sys.time()
  pdat <- t(apply(dat, 1, sample))
  fl <- flash_init(pdat, var_type = 0) |>
    flash_set_verbose(0) |>
    flash_greedy(Kmax = 1L,
                 ebnm_fn = c(flash_ebnm(group = sites_to_keep$chrom), 
                             ebnm_normal)) 
  elbo <- fl$elbo
  fl <- fl |>
    flash_greedy(Kmax = 1L,
                 ebnm_fn = c(flash_ebnm(group = sites_to_keep$chrom), 
                             ebnm_normal)) 
  tlist[i] <- Sys.time() - t0
  pvelist[i] <- fl$pve[2]
  elbodifflist[i] <- fl$elbo - elbo
}
```

#### Timings (in minutes):

```{r}
boxplot(tlist)
```

#### ELBO difference:

```{r}
boxplot(elbodifflist)
```

#### PVE:

```{r}
boxplot(pvelist)
```


## Conclusions

**Results are very consistent across iterations**, which suggests that we only need to do a small number of permutations (maybe two or three is enough)? **The permutation-based PVE cutoff also makes a lot of sense**; surprisingly, it is very much in line with results I obtained using per-chromosome fits (there were 25--30 factors in most of my per-chromosome fits, and the PVE cutoff gives us $K \approx 30$ above).

As far as implementing this approach goes, **I would prefer to add ~5 factors at a time and then check PVE, stopping when at least 2 or 3 do not meet the PVE threshold.** The PVE is not monotonic as a function of $k$, so there is a possibility that we stop too early if we reject the first factor that does not meet our threshold (the `flashier` plots are somewhat deceptive, since they re-arrange the factors to produce a monotonic curve). 
